---
title: "Chapter 1"
author: "Shonda Kuiper and Ankur Roy"
date: "2024-6-20"
output:
  html_document:
    toc: yes
    toc_float:
      collapsed: yes
      smooth_scroll: no
  pdf_document:
    toc:no
  word_document:
    toc: no
---
## **[Randomization Tests: Birdnest case study]{.ul}**

**Installing packages and datasets**

The first step in creating an RMD to submit homework assignments is to require and load the needed packages. These packages are already installed on the online RStudio server, but if you are using your own version of R, you will need to install packages first. You cannot install packages in an RMD file, but in the console you will need to type: install.packages(“package_name”). Notice that the package names needs to be in quotes. The suggested packages and how to load them are shown below:

library('tidyr')

library('ggplot2')

library('dplyr')

library('mosaic')

library('readr')


After the packages are installed, we will load our data. All datasets from the `Practicing Statistics` Textbook are [here](https://github.com/grinnell-statistics/STA310data):


The first dataset which we use called `Birdnest` is found at thsi url:
```{r url,eval= TRUE,echo=TRUE,message=FALSE}
library(tidyr)
library(ggplot2)
library(dplyr)
library(mosaic)
library(readr)

dataurl<- 'https://raw.githubusercontent.com/grinnell-statistics/STA310data/master/Birdnest.csv'
```

With this url, we can read the `Birdnest` data. Let's assign a variable called
`birdnest` to access the data at.
```{r ex13_read, eval = TRUE,echo=TRUE}
birdnest <- read_csv(dataurl)
```


## **[Summary Statistics and Visualization]{.ul}**

Use the names() and dim() functions to get a feel for our data. In the console (not in an RMD) you can type View(birdnest), to view the dataset.

```{r name_dim,eval=TRUE}
names(birdnest)
dim(birdnest)
View(birdnest)
```


**Note**: A bit about the dataset. You will find different columns with information about 99 bird species in the dataset. The parameters we are concerned with in this example here, are `Length` and `Closed`. `Length` refers to the average size of the bird species, while the `Closed` column is a binary variable, with 1 denoting that the nest type for thst bird is classified as Closed and 0 defining that the nest is open type. In the following activities, we will try to apply various visualization techniques like dotplots/boxplots to perform a comparative study between the sizes of the birds having closed and open nests.



### **Dotplots**

We use `ggplot` and its auxiliary function `geom_dotplot` to plot dotplots for the distribution length based on nest type classification. The plots are stacked towards the center here by using the argument `stackdir=center`

```{r ex13_dot, eval = TRUE}

closed <- birdnest$Closed=='1'
open<- birdnest$Closed=='0'


ggplot(data=birdnest, aes(x = as.factor(Closed), y = Length, fill = as.factor(Closed))) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 1,binwidth=2) +
  labs(x = "Closed", y = "Length") +
  scale_fill_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal()
```


Though the dotplots are impressive, a more visually imformative plot are the boxplots. It gives you statistics summary of the distributions. Use `geom_boxplot` within the library ggplot to group by the variable `Closed` and set the fill color parameter to the same as well. Here you can look at the different quantiles, the max and min along with outliers if any.



### **Boxplots**

```{r ex13_box, eval = TRUE}

birdnest$Closed


p <- ggplot(data=birdnest, aes(x=factor(Closed), y=Length,group=(Closed),fill=factor(Closed))) + geom_boxplot(color='black')+labs(x = "Closed", y = "Length") 

p


```




### **Interactive plotting** 

You can also use the mosaic package to work with interactive graphs. Please use the below syntax for the function called `mplot` and then select your options accordingly.
**Note**: `mplot` should be called in the console and not in the RMD script. You will get to see multiple options to plot from and also a gear button in the `Plots` panel. The gear button allows additional options for plot customization.

```{r mosaic, eval=FALSE}
mplot(birdnest)

```


If you want to display the statistics summary numerically, you can use the `favstats` function within `mosaic` package. You need to call it by selecting the desired variable or parameter you want to compute stats summary for followed by the dataset.

```{r summ, eval=TRUE}
favstats(~Length,data=birdnest)
```



### **Conducting Permutation test**

Permutation tests basically use randomization of a data sample to see how the original test statistic (mean/median) between two groups compare against the randomized statistic.

Let's define a function in R called $randtest$ which does this for us. the general syntax for defining a function: function.name <- function(). The variable `reps` is the permutation number, which is how many times will the data be randomized for this statistic.
```{r random, eval=TRUE,echo=TRUE}

randtest <- function(x, y, fun = mean, reps = 10000) {
  n <- length(x)
  m <- length(y)
  data <- c(x, y)
  results <- numeric(reps)
  
  for (i in 1:reps) {
    simtemp <- sample(data)
    results[i] <- fun(simtemp[1:n]) - fun(simtemp[(n+1):(n+m)])
  }
  
  greater.p <- sum(results >= (fun(x) - fun(y))) / reps
  less.p <- sum(results <= (fun(x) - fun(y))) / reps
  test.stat <- abs(fun(x) - fun(y))
  p.values <- c(greater.p, less.p)
  names(p.values) <- c("p.greater.than", "p.less.than")
  
  return(list(results = results, p.values = p.values, test.stat = test.stat))
}

  

```

This R code creates a function to perform a randomization test on two vectors of data, x and y. The output of the function is a list of length three
items:
1.A vector of length reps representing the approximate randomization distribution;
2.Two p-values: “less than” and “greater than” one-sided p-value;
and
3.The test statistic, which by default, is the difference in means between x and y. 

To run this `randtest` function and view a histogram of only the results vector
```{r rand, eval=TRUE,echo=TRUE}
res <- randtest(closed,open)
```




### **Histograms**

Now, if you want to plot the distribution of the differences of the means, histogram is a convenient tool. For this, we will use `geom_histogram` from `ggplot2` package. The syntax is as follows. Please note that the variable we are interested in one of the return values of the `randtest` function.

```{r hist,eval=TRUE,echo=TRUE}
data_df <- data.frame(value=res[[1]])

ggplot(data_df,aes(x=value))+
  geom_histogram(color='black',fill='blue',bins=30)+
  labs(title='Histogram for mean differences',x='Value',y='Count')

```




### **Bootstrapping the data**

Finally, let's try to construct the Bootstrap distribution for the sample mean.
**Note**: Here the Length column in our `Birdnest` dataset is the third column. So, we can access the whole column by the expression $birdnest[,3]$

Here, we do a couple of things. In the $Length$ column of the data, there were some non-numeric NA values which we need to omit. That is why we introduce a new variable called $clean_data$ which will only contain numeric values.

```{r boot,eval=TRUE,echo=TRUE}

clean_data <- unlist(birdnest[, 3])

# If it's supposed to be numeric, ensure conversion
clean_data <- as.numeric(clean_data)

# Remove NA values
clean_data <- na.omit(clean_data)

samp_size <- length(clean_data)
samp_len <- sample(clean_data,samp_size)
reps <- 10000
means <- numeric(reps)
for (i in 1:reps){
  means[i] <- mean(sample(samp_len,samp_size,replace =TRUE))
}
```

Let's plot a histogram for this distribution
```{r boot_hist, eval=TRUE,echo=TRUE}
data_means <- data.frame(value=means)

ggplot(data_means,aes(x=value))+
  geom_histogram(color='black',fill='orange',bins=30)+
  labs(title='Bootstrap distribution of Mean',x='Resample mean',y='Count')
```

We can also calculate the $t$-and percentile confidence intervals at the 95% confidence level for the bootstrap distribution of the mean. Use the following syntax.

95% Bootstrap t Confidence Interval for the mean
```{r con_int_t,eval=TRUE,echo=TRUE}
lower.limit.t<- mean(means)-qt(.975,samp_size-1)*sd(means)
upper.limit.t<- mean(means)+qt(.975,samp_size-1)*sd(means)
```

95% Bootstrap percentile confidence interval for the mean
```{r con_int_p,eval=TRUE,echo=TRUE}
lower.limit.p<- quantile(means,.025)
upper.limit.p<- quantile(means,0.975)
```

